---
title: "PS270 Final Project"
author: "Hallie Haim"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---


## Project thoughts

I am interested in exploring data related to political polarization


I will use the Political Polarization in U.S. Congress (1879–2023) dataset, which compiles congressional ideology measures from Voteview using DW-NOMINATE scores developed by Keith T. Poole and Howard Rosenthal. 
My updated research question is: How does political polarization affect legislative productivity in Congress over time?

hypothesize that higher levels of political polarization lead to fewer laws being passed, because increased ideological division makes compromise and lawmaking more difficult. The key explanatory variable is Polarization (Party.mean.diff.d1).
The dependent variable is number of laws passed per year.

My hypothesis would be supported if the regression shows a negative relationship, meaning higher polarization is associated with fewer laws passed. 


```{r}
library(readr)
library(tidyverse)
library(gapminder)
library(dplyr)

polarization <- read_csv(
  "https://raw.githubusercontent.com/justin-2028/Political-Polarization-in-US-Congress/main/Political%20Polarization%20in%20US%20Congress%20(Assorted).csv"
)
polarization

laws <- read_csv("osfstorage-archive/master.csv")
laws
```
## Cleaning the Data

```{r}
library(dplyr)
library(ggplot2)

clean_data <- polarization |>
  select(congress, year, prop.moderate.d1, party.mean.diff.d1) |>
  filter(!is.na(prop.moderate.d1), !is.na(party.mean.diff.d1))
clean_data
```

```{r}
  clean_data <- clean_data |>
  rename(
    Moderates = prop.moderate.d1,
    Polarization = party.mean.diff.d1
  )
clean_data
```

```{r}
laws_clean <- laws |>
  mutate(
    date_of_passage = as.Date(date_of_passage),
    year = as.integer(format(date_of_passage, "%Y"))
  ) |>
  group_by(year) |>
  summarize(Laws_Passed = n())
laws_clean
```


## Join the Data

```{r}
merged_data <- inner_join(clean_data, laws_clean, by = "year")
merged_data
```



## Visualization

```{r}
ggplot(merged_data, aes(x = Polarization, y = Laws_Passed)) +
  geom_point(alpha = 0.6, color = "dodgerblue") +
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +
  labs(
    title = "Political Polarization vs. Legislative Productivity",
    x = "Party Mean Ideological Difference (Polarization)",
    y = "Number of Laws Passed"
  )
```













