---
title: "PS270 Final Project"
author: "Hallie Haim"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
    css: styles.css
---



## Introduction

# Project Thoughts

I am interested in exploring data related to political polarization. It is a passion of mine and something I am incredibly involved with on campus. I would like to see if my efforts in combating political polarization are actually worth my time. Is polarization causing problems for making change in our nation? Does it need to be addressed?

# Question

Using the Political Polarization in U.S. Congress (1879–2023) dataset, which compiles congressional ideology measures from Voteview using DW-NOMINATE scores developed by Keith T. Poole and Howard Rosenthal and the Libgober, Brian, and Tejas V. Bhat. A Comprehensive Dataset of U.S. Federal Laws. 2023. OSF, https://doi.org/10.17605/OSF.IO/MRGHC data set, I will address the research question: *How does political polarization affect legislative productivity in Congress over time?*

# Hypothesis

I hypothesize that if there are **higher levels of political polarization then fewer laws will be passed**, because increased polarization makes compromise and lawmaking more difficult. As polarization increases, legislators are less willing to compromise across party lines, which makes it more difficult to build coalitions necessary to pass legislation. This leaves the country in gridlock which makes the government ineffective when it comes to implementing change. The  **explanatory variable** is **Polarization (Party.mean.diff.d1)** and the **dependent variable** is **number of laws passed per year**. Legislative productivity is measured as the number of federal laws passed per year, calculated by aggregating individual laws from the Libgober and Bhat dataset. Political polarization is measured using the DW-NOMINATE score difference between political parties, which captures ideological distance based on roll-call voting behavior in Congress. My hypothesis would be supported if the regression shows a statistically significant negative relationship, meaning higher polarization is associated with fewer laws passed. This is an observational study over time, so we are looking at laws being passed each year. I will also have to take confounding variables into account that affect both polarization and the number of laws passed, but we will dive into this even further in my analysis.

## Data Section

# Download the Data

```{r}
library(readr)
library(tidyverse)
library(gapminder)
library(dplyr)

polarization <- read_csv(
  "https://raw.githubusercontent.com/justin-2028/Political-Polarization-in-US-Congress/main/Political%20Polarization%20in%20US%20Congress%20(Assorted).csv"
)
polarization

laws <- read_csv("osfstorage-archive/master.csv")
laws
```
# Cleaning the Data

```{r}
library(dplyr)
library(ggplot2)

clean_data <- polarization |>
  select(congress, year, prop.moderate.d1, party.mean.diff.d1) |>
  filter(!is.na(prop.moderate.d1), !is.na(party.mean.diff.d1))
clean_data
```

```{r}
  clean_data <- clean_data |>
  rename(
    Moderates = prop.moderate.d1,
    Polarization = party.mean.diff.d1)
clean_data
knitr::kable(head(clean_data), caption = "Preview of Polarization Data")
```



```{r}
laws_clean <- laws |>
  mutate(
    date_of_passage = as.Date(date_of_passage),
    year = as.integer(format(date_of_passage, "%Y"))
  ) |>
  group_by(year) |>
  summarize(Laws_Passed = n())
laws_clean
knitr::kable(head(laws_clean), caption = "Preview of Laws Passed Data")
```


# Join the Data

```{r}
merged_data <- inner_join(clean_data, laws_clean, by = "year")
merged_data
knitr::kable(head(merged_data), caption = "Preview of Merged Dataset")
```


# Visualization: Summarizing the Outcome Variable

```{r}
ggplot(laws_clean, aes(x = year, y = Laws_Passed)) +
  geom_line() +
  labs(
    title = "Number of Laws Passed Over Time",
    x = "Year",
    y = "Laws Passed")
```

## Results Section

# Main Graph for Analysis

```{r}
ggplot(merged_data, aes(x = Polarization, y = Laws_Passed)) +
  geom_point(alpha = 0.6, color = "dodgerblue") +
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +
  labs(
    title = "Political Polarization vs. Legislative Productivity",
    x = "Party Mean Ideological Difference (Polarization)",
    y = "Number of Laws Passed")
```
# Linear Regression Model

```{r}
library(modelsummary)
model <- lm(Laws_Passed ~ Polarization, data = merged_data)
modelsummary(model, 
             stars = TRUE, 
             title = "Effect of Political Polarization on Laws Passed")
```
# Interpretation

This graph shows a pretty strong negative relationship between polarization and the number of laws that get passed by congress, which directly supports my hypothesis. When polarization increases by 1 unit, the number of laws passed decreases by about 593 laws. While a one-unit increase represents a large shift in polarization, the negative coefficient indicates that even smaller increases in polarization are associated with meaningful declines in legislative productivity. If polarization didn't exist and it was somehow zero, there would be approximately 626 laws passed per year. 

The extremely small p-value (p < 0.001) gives strong evidence against the null hypothesis of no relationship; and, because the sample size is relatively large, it shows that the relationship is not only statistically significant but also quite large.

It is important to note, however, that while the relationship is strong in direction, there could be confounding variables also influencing how many laws Congress passes in a given period. Correlation does not equal causation; therefore, this should not be interpreted casually. A divided government may both increase polarization and reduce legislative output, making it difficult to isolate the independent effect of polarization alone. There are also partisan shifts in agenda-setting priorities that may also influence both variables. Economic and political crises may also play a role; during periods of war, recession, or national emergencies. Because these factors are not included in the model, the estimated relationship between polarization and laws passed may reflect both the direct effect of polarization and the influence of omitted variables.

## Conclusion

Overall, this analysis finds both the visual and regression results to be strong evidence that as polarization increases, the number of laws passed declines, providing clear support for the hypothesis. The estimated effect is both substantively large and statistically significant as well. Once again, this relationship should not be interpreted as causal because of potential confounding factors such as divided government, different administrations, and major political or economic events. These may influence both polarization and legislative productivity, making it difficult to isolate the independent effect of polarization.

There is a small risk, statistically speaking, of a Type I error in this analysis, meaning we might wrongly say there is a relationship when there isn't. But the risk is very low, the result was statistically significant at the p < 0.001 level. If I had more time or resources, I would consider expanding the analysis to include additional control variables such as party control of Congress, economic indicators, or major national events to better account for confounding variables.
