---
title: "PS270 Final Project"
author: "Hallie Haim"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---


## Project thoughts

I am interested in exploring data related to political polarization


I will use the Political Polarization in U.S. Congress (1879–2023) dataset, which compiles congressional ideology measures from Voteview using DW-NOMINATE scores developed by Keith T. Poole and Howard Rosenthal. 
My tentative research question is: How does the proportion of moderate legislators in Congress affect the level of party polarization over time? 
I hypothesize that if the proportion of moderates declines, then the party polarization increases because when fewer moderates are present the ideological distance between parties increases.
The key explanatory variable is Prop.moderate.d1 (proportion of legislators in each Congress whose DW-NOMINATE scores fall between −0.25 and +0.25).
The dependent variable is Party.mean.diff.d1 (the difference between Democratic and Republican party means indicating polarization).
My hypothesis would be supported if the regression shows a negative relationship meaning lower proportions of moderates are associated with greater party mean differences. 


```{r}
library(readr)

polarization <- read_csv(
  "https://raw.githubusercontent.com/justin-2028/Political-Polarization-in-US-Congress/main/Political%20Polarization%20in%20US%20Congress%20(Assorted).csv"
)
head(polarization)
```
## Cleaning the Data

```{r}
library(dplyr)
library(ggplot2)

clean_data <- polarization |>
  select(congress, prop.moderate.d1, party.mean.diff.d1) |>
  filter(!is.na(prop.moderate.d1), !is.na(party.mean.diff.d1))
clean_data
```

```{r}
  clean_data <- clean_data |>
  rename(
    Moderates = prop.moderate.d1,
    Polarization = party.mean.diff.d1
  )
clean_data
```
## Visualization

```{r}
ggplot(clean_data, aes(x = Moderates, y = Polarization)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +
  labs(
    title = "Moderates vs. Party Polarization in U.S. Congress",
    x = "Proportion of Moderate Legislators",
    y = "Party Mean Ideological Difference"
  )
```













